---
title: "15 Insights from North Carolina School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{15 Insights from North Carolina School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

```{r load-packages}
library(ncschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_minimal(base_size = 14))
```

## 1. North Carolina gained 118,000 students since 2006

One of America's fastest-growing school systems.

```{r statewide-data}
enr <- fetch_enr_multi(c(2006, 2010, 2015, 2020, 2024), use_cache = TRUE)

statewide <- enr %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  select(end_year, n_students)

stopifnot(nrow(statewide) > 0)
statewide
```

```{r statewide-chart}
ggplot(statewide, aes(x = end_year, y = n_students / 1e6)) +
  geom_line(color = "#2171B5", linewidth = 1.2) +
  geom_point(color = "#2171B5", size = 3) +
  geom_text(aes(label = scales::comma(n_students)), vjust = -1, size = 3.5) +
  scale_y_continuous(
    labels = scales::label_number(suffix = "M"),
    limits = c(1.2, 1.7)
  ) +
  labs(
    title = "North Carolina Public School Enrollment",
    x = "Year",
    y = "Total Students"
  )
```

## 2. Wake County is the state's largest district

The Research Triangle's anchor district has nearly 160,000 students.

```{r top-districts-data}
enr_2024 <- fetch_enr(2024, use_cache = TRUE)

top_districts <- enr_2024 %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  arrange(desc(n_students)) %>%
  head(10) %>%
  select(district_name, n_students)

stopifnot(nrow(top_districts) > 0)
top_districts
```

```{r top-districts-chart}
top_districts %>%
  mutate(district_name = reorder(district_name, n_students)) %>%
  ggplot(aes(x = n_students / 1000, y = district_name)) +
  geom_col(fill = "#2171B5") +
  geom_text(aes(label = scales::comma(n_students)), hjust = -0.1, size = 3.5) +
  scale_x_continuous(
    labels = scales::label_number(suffix = "K"),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Top 10 North Carolina School Districts by Enrollment (2024)",
    x = "Students (thousands)",
    y = NULL
  )
```

## 3. Hispanic enrollment surpassed Asian enrollment fivefold

North Carolina's demographic transformation since 2018 is dramatic.

```{r demographics-data}
enr_demo <- fetch_enr_multi(c(2018, 2019, 2020, 2021, 2024), use_cache = TRUE)

demographics <- enr_demo %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian")) %>%
  select(end_year, subgroup, n_students) %>%
  mutate(subgroup = factor(subgroup,
    levels = c("white", "black", "hispanic", "asian"),
    labels = c("White", "Black", "Hispanic", "Asian")))

stopifnot(nrow(demographics) > 0)
demographics
```

```{r demographics-chart}
ggplot(demographics, aes(x = end_year, y = n_students / 1000, color = subgroup)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "gray50", alpha = 0.7) +
  annotate("text", x = 2020.5, y = 780, label = "COVID", color = "gray50", hjust = -0.2, size = 3.5) +
  scale_color_manual(values = c(
    "White" = "#4292C6",
    "Black" = "#807DBA",
    "Hispanic" = "#41AB5D",
    "Asian" = "#EF6548"
  )) +
  scale_y_continuous(labels = scales::label_number(suffix = "K")) +
  labs(
    title = "Demographic Shifts in NC Public Schools (2018-2024)",
    x = "Year",
    y = "Students (thousands)",
    color = "Race/Ethnicity"
  ) +
  theme(legend.position = "bottom")
```

## 4. Charlotte-Mecklenburg lost students post-COVID

Enrollment decline hit North Carolina's second-largest district.

```{r cms-data}
cms_trend <- enr_demo %>%
  filter(is_district, district_id == "600", subgroup == "total_enrollment",
         grade_level == "TOTAL") %>%
  select(end_year, district_name, n_students) %>%
  mutate(change = n_students - lag(n_students))

stopifnot(nrow(cms_trend) > 0)
cms_trend
```

```{r cms-chart}
ggplot(cms_trend, aes(x = end_year, y = n_students / 1000)) +
  geom_line(color = "#CB181D", linewidth = 1.2) +
  geom_point(color = "#CB181D", size = 3) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "gray50", alpha = 0.7) +
  annotate("text", x = 2020.5, y = 150, label = "COVID", color = "gray50", hjust = -0.2, size = 3.5) +
  geom_text(aes(label = scales::comma(n_students)), vjust = -1, size = 3.5) +
  scale_y_continuous(
    labels = scales::label_number(suffix = "K"),
    limits = c(130, 155)
  ) +
  labs(
    title = "Charlotte-Mecklenburg Schools: Post-COVID Enrollment",
    x = "Year",
    y = "Students (thousands)"
  )
```

## 5. Charter schools serve nearly 10% of NC students

North Carolina's 219 charter campuses enroll over 143,000 students.

```{r charter-data}
charter_summary <- enr_2024 %>%
  filter(is_campus, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  group_by(is_charter) %>%
  summarize(
    n_schools = n(),
    students = sum(n_students, na.rm = TRUE),
    .groups = "drop"
  )

state_total <- enr_2024 %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  pull(n_students)

charter_summary <- charter_summary %>%
  mutate(pct = round(students / state_total * 100, 1))

stopifnot(nrow(charter_summary) > 0)
charter_summary
```

```{r charter-chart}
ggplot(charter_summary, aes(x = ifelse(is_charter, "Charter", "Traditional"), y = students / 1000, fill = is_charter)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(scales::comma(students), "\n(", pct, "%)")), vjust = -0.3, size = 4) +
  scale_fill_manual(values = c("FALSE" = "#2171B5", "TRUE" = "#41AB5D"), guide = "none") +
  scale_y_continuous(
    labels = scales::label_number(suffix = "K"),
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title = "Charter vs Traditional School Enrollment (2024)",
    x = NULL,
    y = "Students (thousands)"
  )
```

## 6. The coast is booming while the Piedmont stalls

Brunswick and New Hanover counties are growing; Greensboro-area enrollment is flat.

```{r regional-data}
enr_regional <- fetch_enr_multi(c(2015, 2024), use_cache = TRUE)

coastal <- c("New Hanover", "Brunswick", "Pender")
piedmont <- c("Guilford", "Forsyth", "Alamance")

regional <- enr_regional %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  mutate(region = case_when(
    grepl(paste(coastal, collapse = "|"), district_name) ~ "Coast",
    grepl(paste(piedmont, collapse = "|"), district_name) ~ "Piedmont",
    TRUE ~ "Other"
  )) %>%
  filter(region %in% c("Coast", "Piedmont")) %>%
  group_by(end_year, region) %>%
  summarize(total = sum(n_students, na.rm = TRUE), .groups = "drop")

stopifnot(nrow(regional) > 0)
regional
```

```{r regional-chart}
ggplot(regional, aes(x = factor(end_year), y = total / 1000, fill = region)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(
    aes(label = scales::comma(total)),
    position = position_dodge(width = 0.7),
    vjust = -0.5,
    size = 3.5
  ) +
  scale_fill_manual(values = c("Coast" = "#41AB5D", "Piedmont" = "#EF6548")) +
  scale_y_continuous(
    labels = scales::label_number(suffix = "K"),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Coastal vs Piedmont School Enrollment",
    x = "Year",
    y = "Students (thousands)",
    fill = "Region"
  ) +
  theme(legend.position = "bottom")
```

## 7. Economically disadvantaged students are half of enrollment

Poverty defines North Carolina schools.

```{r econ-disadv-data}
econ_data <- enr_2024 %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("total_enrollment", "econ_disadv")) %>%
  select(subgroup, n_students) %>%
  mutate(pct = round(n_students / max(n_students) * 100, 1))

stopifnot(nrow(econ_data) > 0)
econ_data
```

```{r econ-disadv-chart}
ggplot(econ_data, aes(x = reorder(subgroup, -n_students), y = n_students / 1000, fill = subgroup)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(scales::comma(n_students), "\n(", pct, "%)")), vjust = -0.3, size = 4) +
  scale_fill_manual(
    values = c("total_enrollment" = "#2171B5", "econ_disadv" = "#CB181D"),
    guide = "none"
  ) +
  scale_x_discrete(labels = c("total_enrollment" = "All Students", "econ_disadv" = "Econ. Disadvantaged")) +
  scale_y_continuous(
    labels = scales::label_number(suffix = "K"),
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title = "Half of NC Students Are Economically Disadvantaged (2024)",
    x = NULL,
    y = "Students (thousands)"
  )
```

## 8. Durham demographics are shifting rapidly

The Triangle's most diverse district is transforming.

```{r durham-data}
durham_demographics <- enr_demo %>%
  filter(is_district, district_id == "320", grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian")) %>%
  group_by(end_year) %>%
  mutate(
    total = sum(n_students),
    pct = round(n_students / total * 100, 1)
  ) %>%
  ungroup() %>%
  select(end_year, subgroup, n_students, pct)

stopifnot(nrow(durham_demographics) > 0)
durham_demographics
```

```{r durham-chart}
durham_demographics %>%
  mutate(subgroup = factor(subgroup,
    levels = c("white", "black", "hispanic", "asian"),
    labels = c("White", "Black", "Hispanic", "Asian"))) %>%
  ggplot(aes(x = end_year, y = pct, fill = subgroup)) +
  geom_area(alpha = 0.7) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "white", alpha = 0.7) +
  scale_fill_manual(values = c(
    "White" = "#4292C6",
    "Black" = "#807DBA",
    "Hispanic" = "#41AB5D",
    "Asian" = "#EF6548"
  )) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  labs(
    title = "Durham Public Schools: Demographic Transformation",
    x = "Year",
    y = "Share of Enrollment",
    fill = "Race/Ethnicity"
  ) +
  theme(legend.position = "bottom")
```

## 9. English Learners grew 42% from 2018 to 2024

NC schools are adapting to a multilingual reality, adding nearly 50,000 EL students in six years.

```{r el-data}
el_trend <- enr_demo %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "lep") %>%
  select(end_year, n_students) %>%
  mutate(pct_change = round((n_students / first(n_students) - 1) * 100, 1))

stopifnot(nrow(el_trend) > 0)
el_trend
```

```{r el-chart}
ggplot(el_trend, aes(x = end_year, y = n_students / 1000)) +
  geom_area(fill = "#41AB5D", alpha = 0.3) +
  geom_line(color = "#41AB5D", linewidth = 1.2) +
  geom_point(color = "#41AB5D", size = 3) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "gray50", alpha = 0.7) +
  annotate("text", x = 2020.5, y = 185, label = "COVID", color = "gray50", hjust = -0.2, size = 3.5) +
  geom_text(aes(label = scales::comma(n_students)), vjust = -1, size = 3.5) +
  scale_y_continuous(
    labels = scales::label_number(suffix = "K"),
    limits = c(0, 200)
  ) +
  labs(
    title = "English Learners in North Carolina Schools",
    x = "Year",
    y = "English Learners (thousands)"
  )
```

## 10. Rural eastern NC is losing students fast

Tobacco country is emptying out while cities grow.

```{r eastern-rural-data}
enr_multi <- fetch_enr_multi(c(2015, 2024), use_cache = TRUE)

# Eastern rural counties (traditional tobacco belt)
eastern_rural <- c("Edgecombe", "Halifax", "Hertford", "Northampton",
                   "Bertie", "Martin", "Washington", "Tyrrell")

eastern_data <- enr_multi %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  mutate(is_eastern = grepl(paste(eastern_rural, collapse = "|"), district_name)) %>%
  filter(is_eastern) %>%
  group_by(end_year) %>%
  summarize(total = sum(n_students, na.rm = TRUE), .groups = "drop") %>%
  mutate(pct_change = round((total / first(total) - 1) * 100, 1))

stopifnot(nrow(eastern_data) > 0)
eastern_data
```

```{r eastern-rural-chart}
ggplot(eastern_data, aes(x = factor(end_year), y = total / 1000)) +
  geom_col(fill = "#8B4513", width = 0.6) +
  geom_text(aes(label = scales::comma(total)), vjust = -0.5, size = 4) +
  scale_y_continuous(
    labels = scales::label_number(suffix = "K"),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Eastern NC Rural Counties: Enrollment Decline",
    x = "Year",
    y = "Students (thousands)"
  )
```

## 11. Union County added 10,000 students then plateaued

Charlotte's southern suburbs surged from 31,000 to 41,000 (2006-2015), then leveled off.

```{r union-growth-data}
enr_union <- fetch_enr_multi(c(2006, 2010, 2015, 2020, 2024), use_cache = TRUE)

union_trend <- enr_union %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Union", district_name)) %>%
  select(end_year, district_name, n_students)

stopifnot(nrow(union_trend) > 0)
union_trend
```

```{r union-growth-chart}
ggplot(union_trend, aes(x = end_year, y = n_students / 1000)) +
  geom_area(fill = "#2171B5", alpha = 0.3) +
  geom_line(color = "#2171B5", linewidth = 1.2) +
  geom_point(color = "#2171B5", size = 3) +
  geom_text(aes(label = scales::comma(n_students)), vjust = -1, size = 3.5) +
  scale_y_continuous(
    labels = scales::label_number(suffix = "K"),
    limits = c(0, 50)
  ) +
  labs(
    title = "Union County Schools: Charlotte's Growth Engine",
    x = "Year",
    y = "Students (thousands)"
  )
```

## 12. Asheville's mountain districts saw enrollment decline

Western NC has an aging population and shrinking schools.

```{r mountain-data}
enr_mountain <- fetch_enr_multi(c(2015, 2024), use_cache = TRUE)

# Mountain counties around Asheville
mountain <- c("Buncombe", "Henderson", "Haywood", "Madison",
              "Transylvania", "Yancey", "Mitchell")

mountain_data <- enr_mountain %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  mutate(is_mountain = grepl(paste(mountain, collapse = "|"), district_name)) %>%
  filter(is_mountain) %>%
  group_by(end_year) %>%
  summarize(
    total = sum(n_students, na.rm = TRUE),
    n_districts = n(),
    .groups = "drop"
  )

stopifnot(nrow(mountain_data) > 0)
mountain_data
```

```{r mountain-chart}
ggplot(mountain_data, aes(x = factor(end_year), y = total / 1000)) +
  geom_col(fill = "#4A7C59", width = 0.6) +
  geom_text(aes(label = scales::comma(total)), vjust = -0.5, size = 4) +
  scale_y_continuous(
    labels = scales::label_number(suffix = "K"),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Western NC Mountain Counties: Enrollment Trends",
    x = "Year",
    y = "Students (thousands)"
  )
```

## 13. Special education share held steady at 13-14% since 2018

Despite a 3% headcount decline, special education's share of total enrollment barely moved.

```{r sped-data}
enr_sped <- fetch_enr_multi(c(2018, 2019, 2021, 2024), use_cache = TRUE)

sped_trend <- enr_sped %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("total_enrollment", "special_ed")) %>%
  select(end_year, subgroup, n_students) %>%
  pivot_wider(names_from = subgroup, values_from = n_students) %>%
  mutate(
    pct_sped = round(special_ed / total_enrollment * 100, 1),
    sped_change = round((special_ed / first(special_ed) - 1) * 100, 1)
  )

stopifnot(nrow(sped_trend) > 0)
sped_trend
```

```{r sped-chart}
ggplot(sped_trend, aes(x = end_year)) +
  geom_col(aes(y = special_ed / 1000), fill = "#6A51A3", width = 2) +
  geom_text(aes(y = special_ed / 1000, label = paste0(pct_sped, "%")),
            vjust = -0.5, size = 4) +
  scale_y_continuous(
    labels = scales::label_number(suffix = "K"),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Special Education Students in NC",
    x = "Year",
    y = "Students with IEPs (thousands)"
  )
```

## 14. The Triangle vs Triad: diverging trajectories

Raleigh-Durham grows while Greensboro-Winston stalls.

```{r triangle-triad-data}
enr_metro <- fetch_enr_multi(c(2015, 2020, 2024), use_cache = TRUE)

triangle <- c("Wake", "Durham", "Orange", "Johnston", "Chatham")
triad <- c("Guilford", "Forsyth", "Davidson", "Randolph", "Alamance")

metro_data <- enr_metro %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  mutate(region = case_when(
    grepl(paste(triangle, collapse = "|"), district_name) ~ "Triangle",
    grepl(paste(triad, collapse = "|"), district_name) ~ "Triad",
    TRUE ~ "Other"
  )) %>%
  filter(region %in% c("Triangle", "Triad")) %>%
  group_by(end_year, region) %>%
  summarize(total = sum(n_students, na.rm = TRUE), .groups = "drop")

stopifnot(nrow(metro_data) > 0)
metro_data
```

```{r triangle-triad-chart}
ggplot(metro_data, aes(x = end_year, y = total / 1000, color = region)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_text(aes(label = scales::comma(total)), vjust = -1, size = 3.5, show.legend = FALSE) +
  scale_color_manual(values = c("Triangle" = "#2171B5", "Triad" = "#CB181D")) +
  scale_y_continuous(labels = scales::label_number(suffix = "K")) +
  labs(
    title = "Triangle vs Triad: Diverging Metro Areas",
    x = "Year",
    y = "Students (thousands)",
    color = "Metro Area"
  ) +
  theme(legend.position = "bottom")
```

## 15. COVID erased 56,000 students in a single year

The 2020-2021 school year saw the sharpest enrollment drop in NC history, followed by a partial recovery.

```{r covid-data}
covid_trend <- enr_demo %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  select(end_year, n_students) %>%
  mutate(change = n_students - lag(n_students))

stopifnot(nrow(covid_trend) > 0)
covid_trend
```

```{r covid-chart}
ggplot(covid_trend, aes(x = end_year, y = n_students / 1e6)) +
  geom_line(color = "#2171B5", linewidth = 1.2) +
  geom_point(aes(color = ifelse(end_year == 2021, "COVID", "Normal")), size = 4) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "gray50", alpha = 0.7) +
  annotate("text", x = 2020.5, y = 1.55, label = "COVID", color = "gray50", hjust = -0.2, size = 3.5) +
  geom_text(aes(label = scales::comma(n_students)), vjust = -1, size = 3.5) +
  scale_color_manual(values = c("COVID" = "#CB181D", "Normal" = "#2171B5"), guide = "none") +
  scale_y_continuous(
    labels = scales::label_number(suffix = "M"),
    limits = c(1.35, 1.6)
  ) +
  labs(
    title = "NC Enrollment: COVID Drop and Partial Recovery",
    x = "Year",
    y = "Total Students"
  )
```

## Explore the data yourself

```r
library(ncschooldata)

# Fetch recent years
enr <- fetch_enr_multi(2019:2024, use_cache = TRUE)

# State totals
enr %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL")

# Your district
enr %>%
  filter(grepl("Wake", district_name),
         subgroup == "total_enrollment",
         grade_level == "TOTAL")
```

```{r session-info}
sessionInfo()
```
